---
title: "ggtreeExtra"
author: |
  | Shuangbin Xu and GuangChuang Yu
  | School of Basic Medical Sciences, Southern Medical University
date: "`r Sys.Date()`"
bibliography: ggtreeExtra.bib
biblio-style: apalike
output:
  prettydoc::html_pretty:
    toc: true
    theme: cayman
    highlight: vignette
  pdf_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{ggtreeExtra}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, results="asis", message=FALSE, KnitrSetUp}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

Biocpkg <- function (pkg){
        sprintf("[%s](http://bioconductor.org/packages/%s)", pkg, pkg)
}

CRANpkg <- function(pkg){
        cran <- "https://CRAN.R-project.org/package"
    fmt <- "[%s](%s=%s)"
        sprintf(fmt, pkg, cran, pkg)
}
```

# Setup

```{r, message=FALSE, setup}
library(ggtreeExtra)
library(ggstar)
library(ggplot2)
library(ggtree)
library(treeio)
library(ggnewscale)
```

# 1. Introduction

Phylogenetic trees was often used with associated data in various biological studies. `r Biocpkg("ggtree")`, a flexible R package to visualize phylogenetic tree, had been developed by GuangChuang Yu [@yu2017ggtree]. It provided `geom_facet` function to align associated graphs to the tree [@yu2018two; @yu2020ggtree]. However, This function only works with `rectangular`, `roundrect`, `ellipse` and `slanted` layouts to present tree data on external panels. There is no direct supports in `ggtree` to present associated data on the outer rings of phylogenetic tree in `circular`, `fan` or `radial` layout. To solve the problem, We developed `ggtreeExtra`, which can align associated graphs to `circular`, `fan` or `radial` and other `rectangular` layout tree. `ggtreeExtra` provides function, `geom_fruit` to align graphs to the tree. But the associated graphs will align at the different position of external panels of the tree. So we also developed `geom_fruit_list` to add multiple layers at the same position of external panels of tree. Furthermore, `axis` of external layers can be added using the `axis.params=list(axis="x",...)` in `geom_fruit`. The `grid lines` of external layers can be added using the `grid.params=list()` in `geom_fruit`. These functions are based on `r CRANpkg("ggplot2")` using grammar of graphics [@hadlyggplot2]. More vignettes can be found on the [`chapter10`](https://yulab-smu.top/treedata-book/chapter10.html) of [online book](http://yulab-smu.top/treedata-book).

# 2. Install

You can use the following to install it

```{r, eval=FALSE, INSTALL}
# for devel
if(!requireNamespace("remotes", quietly=TRUE)){
    install.packages("remotes")
}
remotes::install_github("YuLab-SMU/ggtreeExtra")

# for release
if (!requireNamespace("BiocManager", quietly=TRUE))
    install.packages("BiocManager")
    
## BiocManager::install("BiocUpgrade") ## you may need this
BiocManager::install("ggtreeExtra")
```

# 3. Usage

To show the package profiling, I will use a tree file downloaded from [plotTree](https://github.com/katholt/plotTree/blob/master/tree_example_april2015/tree.nwk). The associated datasets were simulated, All of them have one column contained taxa labels of tree. 

## 3.1 add single layer

```{r, fig.align="center", fig.height=7, fig.width=7, warning=FALSE, singlelayer}
# The path of tree file.
trfile <- system.file("extdata", "tree.nwk", package="ggtreeExtra")
# The path of file to plot tip point.
tippoint1 <- system.file("extdata", "tree_tippoint_bar.csv", package="ggtreeExtra")
# The path of first layer outside of tree.
ring1 <- system.file("extdata", "first_ring_discrete.csv", package="ggtreeExtra")
# The path of second layer outside of tree.
ring2 <- system.file("extdata", "second_ring_continuous.csv", package="ggtreeExtra")

# The tree file was imported using read.tree. If you have other tree format files, you can use corresponding functions from treeio package to read it.
tree <- read.tree(trfile)

# This dataset will to be plotted point and bar.
dat1 <- read.csv(tippoint1)
knitr::kable(head(dat1))
# This dataset will to be plotted heatmap
dat2 <- read.csv(ring1)
knitr::kable(head(dat2))
# This dataset will to be plotted heatmap
dat3 <- read.csv(ring2)
knitr::kable(head(dat3))
# The format of the datasets is the long shape for ggplot2. If you have short shape dataset,
# you can use `melt` of `reshape2` or `pivot_longer` of `tidyr` to convert it.

# We use ggtree to create fan layout tree. 
p <- ggtree(tree, layout="fan", open.angle=10, size=0.5)
p

## Next, we can use %<+% of ggtree to add annotation dataset to tree.
#p1 <- p %<+% dat1
#p1
## We use geom_star to add point layer outside of tree.
#p2 <- p1 + 
#      geom_star(
#          mapping=aes(fill=Location, size=Length, starshape=Group),
#          starstroke=0.2
#      ) +
#      scale_size_continuous(
#          range=c(1, 3),
#          guide=guide_legend(
#                     keywidth=0.5, 
#                     keyheight=0.5, 
#                     override.aes=list(starshape=15), 
#                     order=2)
#      ) + 
#      scale_fill_manual(
#          values=c("#F8766D", "#C49A00", "#53B400", "#00C094", "#00B6EB", "#A58AFF", "#FB61D7"),
#          guide="none" # don't show the legend.
#      ) + 
#      scale_starshape_manual(
#          values=c(1, 15),
#          guide=guide_legend(
#                    keywidth=0.5, # adjust width of legend
#                    keyheight=0.5, # adjust height of legend
#                    order=1 # adjust the order of legend for multiple legends.
#                ),
#          na.translate=FALSE # to remove the NA legend.
#      ) 
#p2

# Or if you don't use %<+% to import annotation dataset, instead of `data` parameter of `geom_fruit`. 

# You should specify the column contained tip labels to y axis of `mapping`, here is `y=ID`.

# the `position` parameter was set to `identity` to add the points on the tip nodes.
p2 <- p + 
      geom_fruit(
          data=dat1,
          geom=geom_star,
          mapping=aes(y=ID, fill=Location, size=Length, starshape=Group),
          position="identity",
          starstroke=0.2
      ) + 
      scale_size_continuous(
          range=c(1, 3), # the range of size.
          guide=guide_legend(
                    keywidth=0.5, 
                    keyheight=0.5,
                    override.aes=list(starshape=15),
                    order=2
                )
      ) +
      scale_fill_manual(
          values=c("#F8766D", "#C49A00", "#53B400", "#00C094", "#00B6EB", "#A58AFF", "#FB61D7"),
          guide="none" 
      ) + 
      scale_starshape_manual(
          values=c(1, 15),
          guide=guide_legend(
                    keywidth=0.5,
                    keyheight=0.5,
                    order=1
                )
      )
p2

# Next, we will add a heatmap layer on the outer ring of p2 using `geom_tile` of `ggplot2`.
# Since we want to map some variable of dataset to the `fill` aesthetics of `geom_tile`, but 
# the `fill` of p2 had been mapped. So I need use `new_scale_fill` of `ggnewscale` package to initialize it.
p3 <- p2 + 
      new_scale_fill() + 
      geom_fruit(
          data=dat2,
          geom=geom_tile,
          mapping=aes(y=ID, x=Pos, fill=Type),
          offset=0.08,   # The distance between external layers, default is 0.03 times of x range of tree.
          pwidth=0.25 # width of the external layer, default is 0.2 times of x range of tree.
      ) + 
      scale_fill_manual(
          values=c("#339933", "#dfac03"),
          guide=guide_legend(keywidth=0.5, keyheight=0.5, order=3)
      ) 
p3

# You can also add heatmap layer for continuous values.
p4 <- p3 + 
      new_scale_fill() +
      geom_fruit(
          data=dat3,
          geom=geom_tile,
          mapping=aes(y=ID, x=Type2, alpha=Alpha, fill=Type2),
          pwidth=0.15,
          axis.params=list(
                          axis="x", # add axis text of the layer.
                          text.angle=-45, # the text angle of x-axis.
                          hjust=0  # adjust the horizontal position of text of axis.
                      )
      ) +
      scale_fill_manual(
          values=c("#b22222", "#005500", "#0000be", "#9f1f9f"),
          guide=guide_legend(keywidth=0.5, keyheight=0.5, order=4)
      ) +
      scale_alpha_continuous(
          range=c(0, 0.4), # the range of alpha
          guide=guide_legend(keywidth=0.5, keyheight=0.5, order=5)
      ) 

# Then add a bar layer outside of the tree.
p5 <- p4 + 
      new_scale_fill() +
      geom_fruit(
          data=dat1,
          geom=geom_bar,
          mapping=aes(y=ID, x=Abundance, fill=Location),  #The 'Abundance' of 'dat1' will be mapped to x
          pwidth=0.4,
          stat="identity",
          orientation="y", # the orientation of axis.
          axis.params=list(
                          axis="x", # add axis text of the layer.
                          text.angle=-45, # the text size of axis.
                          hjust=0  # adjust the horizontal position of text of axis.
                      ),
          grid.params=list() # add the grid line of the external bar plot.
      ) + 
      scale_fill_manual(
          values=c("#F8766D", "#C49A00", "#53B400", "#00C094", "#00B6EB", "#A58AFF", "#FB61D7"),
          guide=guide_legend(keywidth=0.5, keyheight=0.5, order=6)
      ) +
      theme(#legend.position=c(0.96, 0.5), # the position of legend.
          legend.background=element_rect(fill=NA), # the background of legend.
          legend.title=element_text(size=7), # the title size of legend.
          legend.text=element_text(size=6), # the text size of legend.
          legend.spacing.y = unit(0.02, "cm")  # the distance of legends (y orientation).
      ) 
p5
```

## 3.2 add multiple layers on the same position.

In the section, we generated a simulated tree and simulated associated datasets. Then, `geom_fruit_list` was used to add multiple layers at the same position of external panels of the tree.

```{r, fig.align="center", fig.width=7, fig.width=7, multilayer}
# To reproduce.
set.seed(1024)
# generate 100 tip point tree.
tr <- rtree(100)
# generate three datasets, which are the same except the third column name.
dt <- data.frame(id=tr$tip.label, value=abs(rnorm(100)), group=c(rep("A",50),rep("B",50)))
df <- dt
dtf <- dt
colnames(df)[[3]] <- "group2"
colnames(dtf)[[3]] <- "group3"
# plot tree 
p <- ggtree(tr, layout="fan", open.angle=0)
p
# the first ring.
p1 <- p + 
      geom_fruit(
          data=dt,
          geom=geom_bar,
          mapping=aes(y=id, x=value, fill=group),
          orientation="y",
          stat="identity"
      ) + 
      new_scale_fill()
p1
# the second ring
# geom_fruit_list is a list, which first element must be layer of geom_fruit.
p2 <- p1 + 
      geom_fruit_list(
          geom_fruit(
              data = df,
              geom = geom_bar,
              mapping = aes(y=id, x=value, fill=group2),
              orientation = "y",
              stat = "identity"
          ),
          scale_fill_manual(values=c("blue", "red")), # To map group2
          new_scale_fill(), # To initialize fill scale.
          geom_fruit(
              data = dt,
              geom = geom_star,
              mapping = aes(y=id, x=value, fill=group),
              size = 2.5,
              color = NA,
              starstroke = 0
          )
      ) + 
      new_scale_fill()
p2

# The third ring
p3 <- p2 + 
      geom_fruit(
          data=dtf,
          geom=geom_bar,
          mapping = aes(y=id, x=value, fill=group3),
          orientation = "y",
          stat = "identity"
      ) +
      scale_fill_manual(values=c("#00AED7", "#009E73"))
p3
```

# 4. Need helps?

If you have questions/issues, please visit [github issue tracker](https://github.com/YuLab-SMU/ggtreeExtra/issues). You also can post to [google group](https://groups.google.com/forum/#!forum/bioc-ggtree). Users are highly recommended to subscribe to the [mailing list](https://groups.google.com/forum/#!forum/bioc-ggtree).

# 5. Session information

Here is the output of sessionInfo() on the system on which this document was compiled:

```{r, echo=FALSE}
sessionInfo()
```

# 6. Reference
